Push-Pull Signal-Function Functional Reactive Programming

### 2012-11-27 Docs/IFL2012/paper.tex
Draft update for IFL paper

\subsection{Goals}
\label{section:System_Design_and_Interface-Goals}

@@

A {\em monad} is a standard, composable abstraction for writing functions with
a context, used in Haskell for IO~\cite{PeytonJones1993,PeytonJones2001} and
other tasks. A monad is simply a 1-arity type constructor together with two
functions. The first function, {\tt return}, takes a value of type {\tt a} and
produces a value of type {\tt m a}. The second, called {\tt bind} and stylized
in the Haskell standard library as the infix operator {\tt (>>=)}, takes a value
of type {\tt m a} and a function of type {\tt a -> m b} and produces a value of
type {\tt m b}. This allows a value to be operated on out of the context and a
new context to be assigned.

A monad can have other primitives which manipulate the context in some way. For 
instance, the primtives in Haskell's {\tt IO} monad produce actions which, when
interpreted as part of the {\tt main} action, produce some side-effect. The
{\tt State} monad provides {\tt get} and {\tt put} operations to work with a 
state value stored in the context.

@@

The evaluation monad is a monad transformer. This permits it to be used in
conjunction with the {\tt IO} monad (or any other monad) to describe how input
is to be obtained for the signal function being evaluated, and how outputs are
to be handled.

@@

-- Pass through the left side of the input unchanged
second :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut)

@@

-- Remove an empty vector on the right
cancelRight :: (sv :^: SVEmpty) :~> sv

-- Add an empty vector on the right
uncancelRight :: sv :~> (sv :^: SVEmpty)

-- Make right-associative
associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))

-- Make left-associative
unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

@@

There are other formulations of a reactive combinator found in FRP literature
and libraries, which may be implemented using the one supplied. These are shown
in Figure~\ref{figure:alternate_switching_combinators} and may be provided in a
future version of the TimeFlies library.

\begin{figure}
\begin{code}
-- Alternate version of switch,
-- implemented in terms of supplied version
switch_gen ::    (svIn :~> (svOut :^: SVEvent a))
              -> (a -> svIn :~> svOut)
              -> svIn :~> svOut
switch_gen sf f =
  switch (sf >>> second (pureEventTransformer f))

-- Supplied version in terms of alternate version
switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
          -> svIn :~> svOut
switch sf = switch_gen sf id

-- Repeated switch, which takes replacement signal functions
-- externally.
rswitch ::    (svIn :~> svOut)
           -> (svIn :^: SVEvent (svIn :~> svOut)) :~> svOut
rswitch sf =
  switch (first sf >>> second (pureEventTransformer rswitch))
\end{code}
\hrule
\caption{Alternate reactivity combinators.}
\label{figure:alternate_switching_combinators}
\end{figure}

@@

-- Apply the function to each input occurrence,
-- and produce an occurrence for each list element
filterList :: (a -> [b]) -> SVEvent a :~> SVEvent b

-- Apply the function to the stored accumulator
-- and the event occurrence, replacing the accumulator
-- and possibly outputting an occurrence
accumulate ::    (a -> b -> (Maybe c, a))
              -> a
              -> SVEvent b :~> SVEvent c

-- Apply the function to the stored accumulator
-- and the event occurrence, replacing the
-- accumulator and outputting an event occurrence
-- for each element of the list
accumulateList ::    (a -> b -> ([c], a))
                  -> a
                  -> SVEvent b :~> SVEvent c
\end{code}
\hrule
\caption{Event-specific combinators.}
\label{figure:event_specific_combinators}
\end{figure}

@@

-- A dummy handler for an empty output
emptyHandler    :: SVHandler out SVEmpty

-- A handler for an updated signal sample
signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)

-- A handler for an event occurrence
eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)

-- Combine handlers for a vector
combineHandlers ::    SVHandler out svLeft
                   -> SVHandler out svRight
                   -> SVHandler out (svLeft :^: svRight)

@@

\begin{figure}
\begin{code}
-- A sample for all leaves of a signal vector
data SVSample sv

-- Create a sample for a signal leaf
sample          :: a -> SVSample (SVSignal a)

-- A dummy sample for an event leaf
sampleEvt       :: SVSample (SVEvent a)

-- A dummy sample for an empty leaf
sampleNothing   :: SVSample SVEmpty

-- Combine two samples
combineSamples  ::    SVSample svLeft
                   -> SVSample svRight
                   -> SVSample (svLeft :^: svRight)
\end{code}
\hrule
\caption{Data type for initial input}
\label{figure:initial_input}
\end{figure}

@@

\begin{figure}
\begin{code}
-- The evaluation monad
data SFEvalT svIn svOut m a
instance MonadTrans (SFEvalT svIn svOut)
instance (Monad m) => Monad (SFEvalT svIn svOut m)
instance (Functor m) => Functor (SFEvalT svIn svOut m)
instance (Monad m, Functor m) => Applicative (SFEvalT svIn svOut m)
instance (MonadIO m) => MonadIO (SVEvalT svIn svOut m)

@@

We therefore start by creating a set of concrete datatypes for the inputs and
outputs of signal functions. These datatypes will be parameterized by the input
and output types of the signal function, and will not be exposed to the user of
the library. Rather, they will specify how data is represented during the
temporal evaluation of signal functions.

We then describe how signal functions are implemented using these concrete
types, along with higher-order functions and continuations.

@@

The types are therefore expressed in the system exactly as they were described
in Section~\ref{section:System_Design_and_Interface}. (To refresh, see
Figure~\ref{figure:signal_vector_types}.) The striking observation about these
types is that they have {\em no data constructors}. There are no values which
take these types.

Instead, we will create concrete representations which are parameterized over
these types. These concrete representations will be expressed as GADTs, allowing
each data constructor of the representation to fill in a specific signal vector
type for the parameter of the representation.

@@

Having specified a datatype for signal functions, we must now provide
combinators which produce signal functions of this type. Each combinator's
implementation must specify how it is initialized, how it samples its input, and
how it responds to event occurrences.

We will not detail every combinator here, but we will discuss each of the
implementation challenges encountered.

As an example of the implementation of combinators, we show the implementation
of the {\tt identity} signal function in Figure~\ref{figure:identity_implementation}.
This signal function simply passes all of its inputs along as outputs. The
initialization function simply passes along the received sample and outputs the
initialized version of the signal function. The initialized version of the input
is similar, but is self-referential. It outputs itself as its replacement. This
is standard for simple and routing combinators which are not reactive, and
simply move samples and event occurrences around.

\begin{figure}
\begin{code}
identity :: sv :~> sv
identity =
  SF (\initSample -> (initSample, identityInit))

identityInit :: SF Initialized sv sv
identityInit =
  SFInit (\dt sigDelta -> (sigDelta, [], identityInit))
         (\evtOcc -> ([evtOcc], identityInit))
\end{code}
\hrule
\caption{Implementation of the {\tt identity} combinator.}
\label{figure:identity_implementation}
\end{figure}

@@

\begin{figure}
\begin{code}
(>>>) ::    (svIn :~> svBetween) 
         -> (svBetween :~> svOut)
         -> (svIn :~> svOut)
(SF sigSampleF1) >>> (SF sigSampleF2) =
  SF (\sigSample -> let (sigSample', sfInit1) = sigSampleF1 sigSample
                        (sigSample'', sfInit2) = sigSampleF2 sigSample'
                    in (sigSample'', composeInit sfInit1 sfInit2))

composeInit ::     SF Initialized svIn svBetween
                -> SF Initialized svBetween svOut
                -> SF Initialized svIn svOut
composeInit (SFInit dtCont1 inputCont1) sf2@(SFInit dtCont2 inputCont2) =
  SFInit
    (\dt sigDelta -> 
       let (sf1MemOutput, sf1EvtOutputs, sf1New) = dtCont1 dt sigDelta
           (sf2MemOutput, sf2EvtOutputs, sf2New) = dtCont2 dt sf1MemOutput
           (sf2EvtEvtOutputs, sf2Newest) = applySF sf2New sf1EvtOutputs
       in (sf2MemOutput,
           sf2EvtOutputs ++ sf2EvtEvtOutputs,
           composeInit sf1New sf2Newest)
    )
    (\evtOcc -> 
      let (sf1Outputs, newSf1) = inputCont1 evtOcc
          (sf2FoldOutputs, newSf2) = applySF sf2 sf1Outputs
      in (sf2FoldOutputs, composeInit newSf1 newSf2)   
    )

applySF ::    SF Initialized svIn svOut
           -> [SVOccurrence svIn]
           -> ([SVOccurrence svOut],
               SF Initialized svIn svOut)
applySF sf indices =
  foldr (\evtOcc (changes, SFInit _ changeCont) ->
           let (newChanges, nextSF) = changeCont evtOcc
               in (newChanges ++ changes, nextSF))
        ([], sf)
        indices
\end{code}
\hrule
\caption{Implementation of serial composition.}
\label{figure:serial_composition_implementation}
\end{figure}

@@

\begin{figure}
\begin{code}
-- T.swap is imported from Data.Tuple
swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
swap =
  SF ((, swapInit) . 
      uncurry combineSamples .
      T.swap . splitSample)

swapInit :: SF Initialized (SVAppend sv1 sv2) (SVAppend sv2 sv1)
swapInit =
  SFInit (flip (const .
                (, [], swapInit) .
                uncurry combineDeltas .
                T.swap . splitDelta))
          (\evtOcc ->
             (case chooseOccurrence evtOcc of
                Left lOcc  -> [occRight lOcc]
                Right rOcc -> [occLeft rOcc], swapInit))

\end{code}
\hrule
\caption{Implementation of the {\tt swap} routing combinator.}
\label{figure:swap_implementation}
\end{figure}

@@

\begin{figure}
\begin{code}
-- | Accumulate over event occurrences
accumulate :: (a -> b -> (Maybe c, a)) -> a -> SVEvent b :~> SVEvent c
accumulate f a = acc a
  where acc a = switch (pureEventTransformer (f a) >>>
                        copy >>>
                        first (pureEventTransformer fst >>> filter id) >>>
                        second (pureEventTransformer (acc . snd)))

-- | Accumulate over event occurrences, with lists of event outputs
accumulateList :: (a -> b -> ([c], a)) -> a -> SVEvent b :~> SVEvent c
accumulateList f a = acc a
  where acc a = switch (pureEventTransformer (f a) >>>
                        copy >>>
                        first (pureEventTransformer fst >>> filterList id) >>>
                        second (pureEventTransformer (acc . snd)))
\end{code}
\hrule
\caption{Implementation of event accumulators.}
\label{figure:accumulate_implementation}
\end{figure}

@@

In a programming language with type constructors, a monad is simply a 1-arity
type constructor together with two operations. We shall call this type
constructor $m$. The first operation takes a value of any type
(call the type $a$) and produces a value of type $m \: a$ ($m$ applied to $a$).
The second operation takes a value of type $m \: a$ and a function from $a$ to
$m \: b$, and produces a $m \: b$. This means that this second operation,
monadic application, applies the function to the value in the context of the
monad.

For our purposes, there are two interesting properties of monads. One is that
for a specific type constructor, opaque primitives may be defined which
can be used, along with return, to build functions for the second argument
of the monadic application operation. In our case, we would like operations
to push an event to a signal function, to update the signal components of
the signal function, and to increment the time and sample the signal function.

This property is what enables monads to be used for input and output. A Haskell
{\tt main} program is an IO action. An IO action is a value of the monadic type
IO. This action is comprised of a primitive action (either a {\tt return} or a 
true IO primitive), possibly together with a function from the output of that
action to another IO action. Note that values with the IO type may appear
anywhere in a Haskell program, but they are only execute when they are returned
as a part of the {\tt main} IO program.

The Haskell typeclass for monads is defined:

\begin{code}
class Monad m where
  return :: a -> m a
  -- | Infix operator for monadic application
  (>>=)  :: m a -> (a -> m b) -> m b 
\end{code}

The other interesting property of monads is the existence of a class of monads
called monad transformers. These are arity-2 type constructors which take as
their first parameter a monad type constructor. The partial application of
such a type constructor to any monad type constructor then results in a monad.
A construct is then provided to lift values of the underlying monad to values
in the transforming monad, which may then be sequenced as normal. This leads
to the powerful concept of monad transformer stacks, where monads with many
different capabilities are combined without any more plumbing than explicit
{\tt lift} operations.

This property will be useful because we want our evaluation interface to be
part of a monad stack, the base of which is Haskell's IO monad. By formulating
the evaluation interface as a monad transformer, we need only define the
operations relevant to the evaluation of signal functions, and we can depend
on the constructs of the IO monad to interact with whatever inputs and outputs
are necessary. In some cases, we may not wish to use the IO monad at all (e.g
for testing or simulation). In this case, we can parameterize over another
monad, such as the Identity monad (which has no special operations and whose
context is just the value), or the State monad (which maintains an implicit 
state accessible by {\tt get} and {\tt put} operations).

The monad tranformer typeclass is defined as:

\begin{code}
class MonadTrans t where
  lift :: Monad m => m a -> t m a
\end{code}

Haskell also supports {\em do}-notation for monads. Do notation simply
makes monadic application implicit between lines and reverses the syntatic
order of lambda bindings, providing a way to use monads in a manner similar
to imperative programming. (See Figure~\ref{figure:monad-example} for an
example.)

\begin{figure}
\begin{code}
-- Do notation:
do putStr "Hello: "
   string <- getStr
   putStrLn $ "Nice to meet you " ++ string
   time <- getCurrentTime
   print time
   putStrLn "Look at the time! Bye."

-- Translates to:
putStr "Hello: " >> getStr >>= 
(\string -> (putStrLn $ "Nice to meet you " ++ string) >>
 getCurrentTime >>=
 (\time -> print time >> putStrLn "Look at the time! Bye."))
\end{code}
\hrule
\caption{An example of monadic $do$-notation and its equivalent monadic
expression.}
\label{figure:monad-example}
\end{figure}

>>>>

@@

@@

The evaluation monad is a monad transformer. This permits it to be used in
conjunction with Haskell's {\tt IO} monad (or any other monad) to describe how
input is to be obtained for the signal function being evaluated, and how outputs
are to be handled.

@@

@@

-- Make right-associative
associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

@@

@@

-- Apply the function to the stored accumulator
-- and the event occurrence, replacing the accumulator
-- and possibly outputting an occurrence
accumulate ::    (a -> b -> (Maybe c, a))
              -> a
              -> SVEvent b :~> SVEvent c
\end{code}
\hrule
\caption{Event-specific combinators.}
\label{figure:event_specific_combinators}
\end{figure}

@@

@@

@@

\begin{figure}
\begin{code}
-- The evaluation monad
data SFEvalT svIn svOut m a
-- Appropriate instances for Monad, MonadTrans, MonadIO,
-- Functor, Applicative, etc

@@

We therefore start by creating a set of concrete datatypes for the inputs and
outputs of signal functions. These datatypes will be parameterized by the input
and output types of the signal function, and will not be exposed to the user of
the library. Rather, they will specify how data is represented during the
temporal evaluation of signal functions. We then implement signal functions
as records of functions from these concrete types to these concrete types paired
with new signal functions. The evaluation system for signal functions then
maintains this record for the signal function it is evaluating, calls the
correct function for the current input, actuates the outputs (using supplied
functions from the outputs to monadic actions) and replaces the record with the
newly supplied signal function record.

@@

We create concrete representations which are parameterized over
these types. These concrete representations will be expressed as GADTs, allowing
each data constructor of the representation to fill in a specific signal vector
type for the parameter of the representation.

@@

Having specified a type constructor for signal functions, we must now provide
combinators which produce signal functions typed with this type constructor.
Space precludes a complete description of the combinators in the TimeFlies
system, but we will present

The simplest example of the implementation of is the {\tt identity} combinator,
shown in Figure~\ref{figure:identity_implementation}. This signal function
simply passes all of its inputs along as outputs. The initialization function
simply passes along the received sample and outputs the initialized version of
the signal function. The initialized version of the input is similar, but is
self-referential. It outputs itself as its replacement. This is standard for
simple and routing combinators which are not reactive, and simply re-arrange,
discard, or combine signals and events.

@@

@@

@@

@@

By formulating
the evaluation interface as a monad transformer, we need only define the
operations relevant to the evaluation of signal functions, and we can depend
on the constructs of the IO monad to interact with whatever inputs and outputs
are necessary. In some cases, we may not wish to use the IO monad at all (e.g
for testing or simulation). In this case, we can parameterize over another
monad, such as the Identity monad (which has no special operations and whose
context is just the value), or the State monad (which maintains an implicit 
state accessible by {\tt get} and {\tt put} operations).
